---
title: "Project 3"
subtitle: "STA6232 - Stats for Data Science II"
author: "Brandy Carr"
format: 
  html:
    grid: 
      sidebar-width: 0px
      body-width: 1400px
      margin-width: 125px
    self-contained: true
    echo: true
    message: false
    warning: false
    error: true
    toc: true
    highlight-style: highlight.theme
editor: source
---

<head>

<link rel="stylesheet" href="https://fonts.googleapis.com/css?family=Montserrat">

<style>

header {
color: #000000 !important;
}

h1, h2, h3, h4 {
color: #1A1A1A !important;
}

b, strong {
color: #3A3A3A !important;
}

body {
color: #7C6FBF !important;
font-family: Montserrat;
}

pre.sourceCode {
background-color: #F9F9F9 !important;
border: 1px solid #CDC8B1;
}

pre>code.sourceCode {
color: #3B3A3A !important;
background-color: #F9F9F9 !important;
font-size: 14px !important;
font-weight: 300 !important;
}

pre>code, div.sourceCode {
font-size: 14px !important;
font-weight: 300 !important;
color: #9A9A8A !important;
background-color: #FFFFFF !important;
} 

code {
color: #6D5FB0 !important;
background-color: #E5E9FF !important;
font-size: 14px !important;
font-weight: 600 !important;
padding: 0.5em !important;
}

</style>

</head>

</br>

**This week you again analyze data from the Jackson Heart Study (JHS). You can find the data on Canvas, in the Week 1 module. For full credit, you must include all code chunks and R output backing up your responses.**

</br>

</br>

## 0

**Import the JHS data.**

```{r}
library(tidyverse)
JHS <- haven::read_sas("analysis1.sas7bdat")
```

</br>

</br>

## 1

**1A.  Model diabetic status (*Diabetes*; 0=non-diabetic, 1=diabetic) as a function of age (*age*; years), weight (*weight*; kg), hypertension status (*HTN*; 0=normotensive, 1=hypertensive), and fasting HDL categorization (*hdl3cat*; 0=low, 1=normal, 2=high), and the interactions between (1) weight and hypertension and (2) weight and fasting HDL categorization. Remember to report the resulting model.**

```{r}
# subset data & removed incomplete cases
JHS1 <- JHS %>% select(Diabetes, age, weight, HTN, hdl3cat) %>% na.omit()

# factored variable hdl3cat
JHS1$HDL <- factor(JHS1$hdl3cat, levels=c(0,1,2), labels=c("low","norm","high"))

# full model
m1 <- glm(Diabetes ~ age + HDL + HTN + weight + weight:HTN + weight:HDL, data=JHS1, family="binomial"(link="logit"))

# reduced model (intercept only)
reduced <- glm(Diabetes ~ 1, data=JHS1, family="binomial"(link="logit"))

# tests significance of the m1 regression line
anova(reduced, m1, test="LRT")

# tests significance of each predictor
car::Anova(m1, type=3)

# tests significance of all terms (pairwise)
summary(m1)
```

`FULL MODEL`

<div style="color: #6D5FB0; background-color: #E5E9FF; border: 1px solid #9083D3; padding: 10px; margin: 0px;">

$$\small{ln \left( \frac{\pi}{1-\pi} \right) = -7.4 + 0.04(age) - 0.6(HDL_{norm}) + 0.36(HDL_{high}) + 2.12(HTN) + 0.03(weight) - 0.01(HTN:weight) + 0.002(HDL_{norm}:weight) - 0.01(HDL_{high}:weight)}$$ {#eq-1A}

</div>

</br>

`COMMENTS`

The full model predicts Diabetes significantly better than the intercept only (reduced) model.

HDL is NOT a significant predictor of Diabetes & neither is it's interaction with weight.

</br>

</br>

**1B.  Perform the appropriate hypothesis test to determine if the interaction between weight and hypertension status is significant. Test at the $\alpha=0.05$ level.**

```{r}
# binary X continuous
round(summary(m1)$coeff[drop=F, "HTN:weight", c(3,4)], 4)
```

`Hypotheses`

$H_{0}$ : $\beta_{HTN:weight} = 0$

$H_{1}$ : $\beta_{HTN:weight} \ne 0$

`Test Statistic`  

$z_{0} = -2.0769$

`P-Value`  

$p < 0.0378$

`Rejection Region`  

Reject $H_{0}$ if $p < \alpha;$ $\phantom0 \alpha = 0.05$

`Conclusion`  

Reject $H_{0}.$

There is sufficient evidence to suggest that the interaction between hypertension status and weight (*HTN:weight*) significantly predicts *Diabetes*.

</br>

</br>

**1C.  Perform the appropriate hypothesis test to determine if the interaction between fasting HDL categorization and weight (*HDL:weight*)is significant. Test at the $\alpha=0.05$ level.**

```{r}
# categotical(c=3) X continuous
round(car::Anova(m1, type=3)[drop=F, "HDL:weight",], 4)
```

`Hypotheses`

$H_{0}$ : $\beta_{HDL_{norm}:weight} = \beta_{HDL_{high}:weight} = 0$

$H_{1}$ : at least one $\beta_{i} \ne 0$

`Test Statistic`  

$\chi^2 = 2.9360$

`P-Value`  

$p = 0.2304$

`Rejection Region`  

Reject $H_{0}$ if $p < \alpha;$ $\phantom0 \alpha = 0.05$

`Conclusion`  

Fail to reject $H_{0}.$

There is NOT sufficient evidence to suggest that the relationship between *Diabetes* and *weight* depends on fasting HDL categorization (*HDL*).

</br>

</br>

**1D.  Create the following models (i.e., plug in the following values and algebraically simplify): (1) weight of 70 kg, (2) weight of 90 kg, and (3) weight of 110 kg.  Remember to report the resulting models.**

$\small{\text{***Displaying rounded coefficients in models below, but used un-rounded values for calculations on the back end}}$

`weight = 70kg`

$$\small{ln \left( \frac{\pi}{1-\pi} \right) = -7.4 + 0.04(age) - 0.6(HDL_{norm}) + 0.36(HDL_{high}) + 2.12(HTN) + 0.03(70) - 0.01(HTN:70) + 0.002(HDL_{norm}:70) - 0.01(HDL_{high}:70)}$$ 

$$\small{ln \left( \frac{\pi}{1-\pi} \right) = -7.4 + 0.04(age) - 0.6(HDL_{norm}) + 0.36(HDL_{high}) + 2.12(HTN) + 2.25 - 0.82(HTN) + 0.15(HDL_{norm}) - 0.69(HDL_{high})}$$

$$\small{ln \left( \frac{\pi}{1-\pi} \right) = -5.155 + 0.042(age) - 0.455(HDL_{norm}) - 0.310(HDL_{high}) + 1.300(HTN)}$$ {#eq-1D-1}

`weight = 90kg`

$$\small{ln \left( \frac{\pi}{1-\pi} \right) = -7.4 + 0.04(age) - 0.6(HDL_{norm}) + 0.36(HDL_{high}) + 2.12(HTN) + 0.03(90) - 0.01(HTN:90) + 0.002(HDL_{norm}:90) - 0.01(HDL_{high}:90)}$$

$$\small{ln \left( \frac{\pi}{1-\pi} \right) = -7.4 + 0.04(age) - 0.6(HDL_{norm}) + 0.36(HDL_{high}) + 2.12(HTN) + 2.89 - 1.06(HTN) + 0.19(HDL_{norm}) - 0.88(HDL_{high})}$$

$$\small{ln \left( \frac{\pi}{1-\pi} \right) = -4.512 + 0.042(age) - 0.414(HDL_{norm}) - 0.506(HDL_{high}) + 1.065(HTN)}$$ {#eq-1D-2}

`weight = 110kg`

$$\small{ln \left( \frac{\pi}{1-\pi} \right) = -7.4 + 0.04(age) - 0.6(HDL_{norm}) + 0.36(HDL_{high}) + 2.12(HTN) + 0.03(110) - 0.01(HTN:110) + 0.002(HDL_{norm}:110) - 0.01(HDL_{high}:110)}$$ 

$$\small{ln \left( \frac{\pi}{1-\pi} \right) = -7.4 + 0.04(age) - 0.6(HDL_{norm}) + 0.36(HDL_{high}) + 2.12(HTN) + 3.54 - 1.29(HTN) + 0.23(HDL_{norm}) - 1.08(HDL_{high})}$$

$$\small{ln \left( \frac{\pi}{1-\pi} \right) = -3.870 + 0.042(age) - 0.372(HDL_{norm}) - 0.702(HDL_{high}) + 0.831(HTN)}$$ {#eq-1D-3}

</br>

</br>

**1E.  Find the odds ratios for the models in 1d.**

```{r}
Models = data.frame(weight_70kg = c(-5.115, 0.042, -0.455, -0.310, 1.300),
                    weight_90kg = c(-4.512, 0.042, -0.414, -0.506, 1.065),
                    weight_110kg = c(-3.870, 0.042, -0.372, -0.702, 0.831))
rownames(Models) = c("Intercept", "age", "HDLnorm", "HDLhigh", "HTN")
Models                
```

```{r}
# Model 1: weight = 70kg
Models[-1,] %>% 
  select(weight_70kg) %>% 
  mutate(OddsRatio = round(exp(weight_70kg), 1), 
         "OddsRatio%" = paste0(round(100*(exp(weight_70kg)-1), 1), "%"))
```

```{r}
# Model 2: weight = 90kg
Models[-1,] %>% 
  select(weight_90kg) %>% 
  mutate(OddsRatio = round(exp(weight_90kg), 1), 
         "OddsRatio%" = paste0(round(100*(exp(weight_90kg)-1), 1), "%"))
```

```{r}
# Model 3: weight = 110kg
Models[-1,] %>% 
  select(weight_110kg) %>% 
  mutate(OddsRatio = round(exp(weight_110kg), 1), 
         "OddsRatio%" = paste0(round(100*(exp(weight_110kg)-1), 1), "%"))
```

</br>

</br>

**1F.  Provide brief interpretations of the odds ratios found in one model (your choice) in 1e.**

Regardless of weight:

- For every 1 year increase in age, the odds of having Diabetes increases by 4.3%.

</br>

For participants who weighed 70 kg (see @eq-1D-1 for the model):

- The odds of having Diabetes is 3.7 times higher for hypertensive participants than it is for normotensive participants.
- As compared to a fasting HDL categorization of low:
  - The odds of having Diabetes for a fasting HDL categorization of normal are decreased by 36.6%.
  - The odds of having Diabetes for a fasting HDL categorization of high are decreased by 26.7%.

</br>

For participants who weighed 90 kg (see @eq-1D-2 for the model):

- The odds of having Diabetes is 2.9 times higher for hypertensive participants than it is for normotensive participants.
- As compared to a fasting HDL categorization of low:
  - The odds of having Diabetes for a fasting HDL categorization of normal are decreased by 33.9%.
  - The odds of having Diabetes for a fasting HDL categorization of high are decreased by 39.7%.

</br>

For participants who weighed 110 kg (see @eq-1D-3 for the model):

- The odds of having Diabetes is 2.3 times higher for hypertensive participants than it is for normotensive participants.
- As compared to a fasting HDL categorization of low:
  - The odds of having Diabetes for a fasting HDL categorization of normal are decreased by 31.1%.
  - The odds of having Diabetes for a fasting HDL categorization of high are decreased by 50.4%.
  
</br>

</br>

**1G.  Construct an appropriate data visualization to help with explaining the model results. Diabetic status should be on the y-axis, age should be on the x-axis, and use the regression lines constructed in 1d.**

```{r}
#| eval: false
#| fig-align: left
#| out-width: 90%
#| fig-asp: 0.5
#| fig-dpi: 300
library(fastDummies)
JHS1 <- JHS1 %>%
  dummy_cols(select_columns = c("HDL")) %>%
  mutate(p_70kg = exp(Models[1,1] + Models[2,1]*age + Models[3,1]*HDL_norm + Models[4,1]*HDL_high + Models[5,1]*HTN) / 
           (1 + exp(Models[1,1] + Models[2,1]*age + Models[3,1]*HDL_norm + Models[4,1]*HDL_high + Models[5,1]*HTN)),
         p_90kg = exp(Models[1,2] + Models[2,2]*age + Models[3,2]*HDL_norm + Models[4,2]*HDL_high + Models[5,2]*HTN) / 
           (1 + exp(Models[1,2] + Models[2,2]*age + Models[3,2]*HDL_norm + Models[4,2]*HDL_high + Models[5,2]*HTN)),
         p_110kg = exp(Models[1,3] + Models[2,3]*age + Models[3,3]*HDL_norm + Models[4,3]*HDL_high + Models[5,3]*HTN) / 
           (1 + exp(Models[1,3] + Models[2,3]*age + Models[3,3]*HDL_norm + Models[4,3]*HDL_high + Models[5,3]*HTN)))

JHS1$HTN2 <- factor(JHS1$HTN, levels=c(0,1), labels=c("Normotensive", "Hypertensive"))
JHS1$HDL2 <- factor(JHS1$hdl3cat, levels=c(0,1,2), labels=c("Low HDL", "Normal HDL", "High HDL"))

JHS1 %>% 
  group_by(HTN2, HDL2) %>% 
  ggplot(aes(x=age, y=Diabetes)) + 
  geom_point(alpha=0.2, size=1) +
  geom_line(aes(y=p_70kg), color = "#00BBA0", linewidth = 0.6, linetype = 1) +
  geom_line(aes(y=p_90kg), color = "#FA962D", linewidth = 0.6, linetype = 1) +
  geom_line(aes(y=p_110kg), color = "#C71585", linewidth = 0.6, linetype = 1) +
  labs(x = "Age", y = "Probability of Diabetes") +
  facet_grid(HTN2 ~ HDL2) +
  theme_bw(base_size = 5.5) +
  theme(title=element_text(face="bold"), legend.position="none")
```

<div style="color: #C71585; background-color: #FEEAF2; border: 1px solid #C71585; padding: 0px; margin-left: 50px; margin-right: 175px; margin-top: 2px; margin-bottom: 2px;"> $$\large{\text{weight = 110 kg}}$$ </div>

<div style="color: #FA962D; background-color: #FFF5F0; border: 1px solid #FA962D; padding: 0px; margin-left: 50px; margin-right: 175px; margin-top: 2px; margin-bottom: 2px;"> $$\large{\text{weight = 90 kg}}$$ </div>

<div style="color: #00BBA0; background-color: #E9FFF9; border: 1px solid #00BBA0; padding: 0px; margin-left: 50px; margin-right: 175px; margin-top: 2px; margin-bottom: 2px;"> $$\large{\text{weight = 70 kg}}$$ </div>

```{r}
#| echo: false
#| fig-align: left
#| out-width: 90%
#| fig-asp: 0.5
#| fig-dpi: 300
library(fastDummies)
JHS1 <- JHS1 %>%
  dummy_cols(select_columns = c("HDL")) %>%
  mutate(p_70kg = exp(Models[1,1] + Models[2,1]*age + Models[3,1]*HDL_norm + Models[4,1]*HDL_high + Models[5,1]*HTN) / 
           (1 + exp(Models[1,1] + Models[2,1]*age + Models[3,1]*HDL_norm + Models[4,1]*HDL_high + Models[5,1]*HTN)),
         p_90kg = exp(Models[1,2] + Models[2,2]*age + Models[3,2]*HDL_norm + Models[4,2]*HDL_high + Models[5,2]*HTN) / 
           (1 + exp(Models[1,2] + Models[2,2]*age + Models[3,2]*HDL_norm + Models[4,2]*HDL_high + Models[5,2]*HTN)),
         p_110kg = exp(Models[1,3] + Models[2,3]*age + Models[3,3]*HDL_norm + Models[4,3]*HDL_high + Models[5,3]*HTN) / 
           (1 + exp(Models[1,3] + Models[2,3]*age + Models[3,3]*HDL_norm + Models[4,3]*HDL_high + Models[5,3]*HTN)))

JHS1$HTN2 <- factor(JHS1$HTN, levels=c(0,1), labels=c("Normotensive", "Hypertensive"))
JHS1$HDL2 <- factor(JHS1$hdl3cat, levels=c(0,1,2), labels=c("Low HDL", "Normal HDL", "High HDL"))

JHS1 %>% 
  group_by(HTN2, HDL2) %>% 
  ggplot(aes(x=age, y=Diabetes)) + 
  geom_point(alpha=0.2, size=1) +
  geom_line(aes(y=p_70kg), color = "#00BBA0", linewidth = 0.6, linetype = 1) +
  geom_line(aes(y=p_90kg), color = "#FA962D", linewidth = 0.6, linetype = 1) +
  geom_line(aes(y=p_110kg), color = "#C71585", linewidth = 0.6, linetype = 1) +
  labs(x = "Age", y = "Probability of Diabetes") +
  facet_grid(HTN2 ~ HDL2) +
  theme_bw(base_size = 5.5) +
  theme(title=element_text(face="bold"), legend.position="none")
```

</br>

</br>

## 2

**2A.  Using ordinal logistic regression, model diabetic status (*diab3cat*; 0=non-diabetic, 1=pre-diabetic, 2=diabetic) as a function of age (*age*; years), weight (*weight*; kg), hypertension status (*HTN*; 0=normotensive, 1=hypertensive), and fasting HDL categorization (*hdl3cat*; 0=low, 1=normal, 2=high). Remember to report the resulting model.**

```{r}
# subset data & removed incomplete cases
JHS2 <- JHS %>% select(diab3cat, age, weight, HTN, hdl3cat) %>% na.omit()

# factored variables diab3cat, HTN, & hdl3cat
JHS2$Diabetes <- factor(JHS2$diab3cat, levels=c(0,1,2), labels=c("Non-Diabetic","Pre-Diabetic","Diabetic"))
JHS2$HTN <- factor(JHS2$HTN, levels=c(0,1), labels=c("Normotensive","Hypertensive"))
JHS2$HDL <- factor(JHS2$hdl3cat, levels=c(0,1,2), labels=c("Low","Normal","High"))

# ordinal logistic regression model
m2 <- MASS::polr(Diabetes ~ age + HDL + HTN + weight, data=JHS2, Hess=T)
summary(m2)
```

`MODELS`

logit$(P[Y\le \text{Non-Diabetic}])$ = 4.995 + 0.051(age) - 0.296(HDLNormal) - 0.587(HDLHigh) + 0.632(Hypertensive) + 0.025(weight)

logit$(P[Y\le \text{Pre-Diabetic}])$ = 6.960 + 0.051(age) - 0.296(HDLNormal) - 0.587(HDLHigh) + 0.632(Hypertensive) + 0.025(weight)

</br>

</br>

**2B.  Which predictors, if any, are significant predictors of diabetic status?**

```{r}
car::Anova(m2, type = 3)
```

All predictors are significant predictors of diabetic status.

</br>

</br>

**2C.  Find the odds ratios and corresponding 95% CIs for the predictors.**

```{r}
OR = round(exp(coefficients(m2)), 2)
cbind(OR, round(exp(confint(m2)), 2))
```

</br>

</br>

**2D.  Provide brief interpretations of the odds ratios found in 2c.**

`age`

For a one year increase in $\text{age}$, the odds in favor of those who are $\text{Pre-Diabetic}$ or lower, as compared to those who are $\text{Diabetic}$, are increased by [$100*(e^{0.0514}-1)$] = 5.3%.

`HDL`

For any fixed response, the estimated odds of those with $\text{Normal HDL}$ is in the $\text{Non-Diabetic}$ direction rather than the $\text{Diabetic}$ direction are [$e^{-0.2963}$] = 0.74 times the estimated odds of those with $\text{Low HDL}$. This is a 26% decrease in odds as compared to those with $\text{Low HDL}$.

For any fixed response, the estimated odds of those with $\text{High HDL}$ is in the $\text{Non-Diabetic}$ direction rather than the $\text{Diabetic}$ direction are [$e^{-0.5873}$] = 0.56 times the estimated odds of those with $\text{Low HDL}$. This is a 44% decrease in odds as compared to those with $\text{Low HDL}$.

`HTN`

For any fixed response, the estimated odds of those with $\text{Hypertension}$ is in the $\text{Non-Diabetic}$ direction rather than the $\text{Diabetic}$ direction are [$e^{0.6321}$] = 1.88 times the estimated odds of those who are $\text{Normotensive}$. This is an 88% increase in odds as compared to those who are $\text{Normotensive}$.

`weight`

For a one lb increase in $\text{weight}$, the odds in favor of those who are $\text{Pre-Diabetic}$ or lower, as compared to those who are $\text{Diabetic}$, are increased by [$100*(e^{0.0246}-1)$] = 2.5%.

</br>

</br>

## 3

**3A.  Using nominal/multinomial logistic regression, model diabetic status (*diab3cat*; 0=non-diabetic, 1=pre-diabetic, 2=diabetic) as a function of age (*age*; years), weight (*weight*; kg), hypertension status (*HTN*; 0=normotensive, 1=hypertensive), and fasting HDL categorization (*hdl3cat*; 0=low, 1=normal, 2=high). Remember to report the resulting model.**

```{r}

```

</br>

**3B.  Which predictors, if any, are significant predictors of diabetic status?**

```{r}

```

</br>

**3C.  Find the odds ratios and corresponding 95% CIs for the predictors.**

```{r}

```

</br>

**3D.  Provide brief interpretations of the odds ratios found in 3c.**

</br>

## 4

**4A.  Consider the proportional odds assumption for ordinal logistic regression. Explain the proportional odds assumption to your supervisor, who is not a statistician or data scientist.**

</br>

**4B. Use the appropriate hypothesis test to determine if we should use model 2a or model 3a when reporting on diabetic status in terms of non-diabetic, pre-diabetic, and diabetic.**

```{r}

```

</br>

**4 Extra credit! Construct a data visualization for the model chosen in 4b to help explain the model's results.**

```{r}

```

</br>

## 5

**Required for graduate students / extra credit for undergraduate students: Write a paragraph to summarize the analyses in this project, written such that a non-quantitative person could understand.**
